---
title: "Reproducible Research Project 1"
author: "Joe Lindenmayer"
date: "December 27, 2020"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
```


First we need to load the data

```{r, echo=TRUE}
archiveFile <- "repdata_data_activity.zip"


if(!file.exists(archiveFile)) {
    
    archiveURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    
    download.file(url=archiveURL,destfile=archiveFile)
}


activity <- read.csv("activity.csv")

activity$date <- ymd(activity$date)
```

Using the data, we want to see the total number of steps taken per day. The results will be plotted in a histogram. For practice, I made the plot using both the base system and ggplot2's qplot function. 

```{r, echo=TRUE}
SPD <- aggregate(steps~date,activity,sum)
hist(SPD$steps,col="lightblue",breaks=20,xlab="Total Steps in a Day",main="Frequency of Total Steps in a Day From October to December")
qplot(SPD$steps,geom="histogram",bins=20,main="Frequency of Total Steps/Day from Oct to Dec",xlab="Total Steps",ylab="Frequency",fill=I("lightblue"),col=I("orange"),alpha=I(0.9))
```

Next, we find the mean and median of the steps taken 

```{r, echo=TRUE}
mean(SPD$steps)
median(SPD$steps)
```

Next, we want to see what activity looks like throughout the day. Data was collected in 5-minute intervals, so we will use this data to produce a time-series line graph to display the average activity throuhgout the day. Once this data is plotted, we also want to find which time interval has the most activity,

```{r, echo=TRUE}
SPI <- aggregate(steps~interval,activity,mean)
with(SPI, plot(interval,steps,type="l"))
ggplot(data=SPI,aes(x=interval,y=steps,group=1)) +theme_minimal()+geom_line(color="orange")+geom_point()
SPI[which.max(SPI[,2]),1]
```


Next, we want to address the several NA entries to the data. We first want to see how many NA entries there are. As a basic solution, we will replace each NA with the average number of steps in a 5-minute interval.

The resulting dataset is then plotted in histogram form, and again for practice, both ggplot and base are used. Finally, the new mean and median with NAs replaced are reported.

```{r, echo=TRUE}
NAIndex <- is.na(activity[,1])
sum(is.na(activity$steps))

averageSteps <- mean(SPI$steps)

activityNEW <- activity
activityNEW[NAIndex,1] <- averageSteps

totalStepsByDay <- aggregate(steps~date,activityNEW,sum)
hist(totalStepsByDay$steps, xlab="Total Steps per Day", ylab="Number of Days Steps were Met",main= "Number of Steps Taken in a Day with NAs Removed")
qplot(totalStepsByDay$steps,geom="histogram",bins=10,xlab="Total Steps Taken",main="Number of Steps Taken per Day with NAs Removed", fill=I("lightblue"),col=I("orange"),alpha=I(0.9))

meanAfterNAs <- mean(totalStepsByDay$steps)
medianAfterNAs <- median(totalStepsByDay$steps)
```

Finally, we would like to see the difference between weekday and weekend activity. Dates were formatted earlier using lubridate, and we will add a new column to the dataset called "datetype" where each interval is assigned weekend or weekday. 

The results are plotted in a line graph. The first plot shows both weekends and weekdays on the same plot, while the second plot displays the results seperately. 

```{r, echo=TRUE}
activity$datetype <- sapply(activity$date, function(x){
    if (weekdays(x) =="Sunday" | weekdays(x) == "Saturday")
        {y <- "Weekend"}
    else{y <- "Weekday"}
    y
    })

weekday_weekend_comparison <- aggregate(steps~interval + datetype, activity, mean, na.rm=TRUE)
plot <- ggplot(weekday_weekend_comparison, aes(x=interval, y=steps, color=datetype))+geom_line() + labs(title = "Average Daily Steps Weekdays Vs Weekends", x="Interval", y="Average # of Steps")
plot1 <- ggplot(weekday_weekend_comparison, aes(x=interval, y=steps, color=datetype))+geom_line() + labs(title = "Average Daily Steps Weekdays Vs Weekends", x="Interval", y="Average # of Steps") + facet_wrap(~datetype, ncol=1,nrow=2)
print(plot)
print(plot1)
```

